Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-05-25 solutions.Rmd
Revise solutions section
Revise to incorporate BJC's comments on the March 14 draft, remove the
experimental-recommendations section, and overhaul the two sections on
absolute DA.

The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.

## Use ratio-based relative-abundance analyses

Since bias creates a consistent fold error in species ratios, a natural approach to countering bias when analyzing relative abundances is to use DA methods that are based on fold changes in ratios.
A variety of such methods have been developed for microbiome DA analysis that build on earlier work from the field of Compositional Data Analysis (CoDA).
These methods analyze log-fold changes in the ratios among a pair of species or, more generally, some product of exponential power of multiple species, which likewise are bias-invariant.
Besides a greater robustness to taxonomic bias, such ratio-based analyses have other advantages.
First, analysis of ratios avoids other standard criticisms leveled at proportion-based analysis—namely that the change in proportion of a species is affected by change in the density of all other species and the set of species used in the denominator of the proportion calculation (@gloor2017micr).
Second, ratios may be more informative than absolute density in some cases, such as for swabs or other sample types for which the absolute density in the sample may be an unreliable indicator of the absolute density of the sampled ecosystem.

Ratios come with their limitations, however.
First, there are a vast number of species ratios (and derivatives) to potentially consider, none of which provide direct information about absolute densities, which can make choosing and interpreting a ratio-based DA analysis challenging.
Second, ratio measurements are impacted by noise in both the numerator and denominator and so can be noisier than proportion measurements, particularly when the read counts in the denominator are small.
A related problem is that, due to the discrete nature of biological organisms and the reads that are generated during sequencing, species in the denominator are commonly observed to have 0 reads; the assumptions that made by different methods about the meaning of these 0 counts can have a major impact on DA results.
Finally, bias invariance at the level of ratios among species does not extend to ratios of additive aggregates of species, such as between bacterial phyla, unless further conditions are met (@mclaren2019cons).

@@

Measurement of community calibration controls along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis.
Community calibration controls are samples whose species identities and relative abundances are known either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
MGS measurement of one or more control communities can be used to directly measure the relative efficiencies among the superset of species in the controls (@mclaren2019cons).
The measured relative efficiencies can then be used to calibrate (remove the effect of bias from) the relative abundances (ratios) of the control species in a set of primary (non-control) experimental samples that were measured with the same protocol (ideally in the same experiment and sequencing run) (@mclaren2019cons).
Calibration can be extended to species not in the controls using statistical methods to impute (predict) the unobserved efficiencies using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).

The calibrated compositions can be used for arbitrary downstream microbiome analyses, including relative and absolute DA analysis.
To demonstrate the potential for calibration to improve fold-change measurements, we estimated bias from one sample in the mock community data from @brooks2015thet that contained all 7 species and used it to calibrate the compositions of all samples (Figure \@ref(fig:calibration-example)).
We then estimated species' densities using the uncalibrated and calibrated proportions using total-abundance normalization, treating the total density as known and fixed.
Visual inspection shows that the bias estimated from just a single control community with all species can be sufficient to greatly reduce the effect of taxonomic bias on the measured fold change in species' proportions and densities (Figure \@ref(fig:calibration-example)).
To demonstrate a practical application, we used the bias estimated from the full set of mock communities as the basis for calibrating the vaginal community time series measurements from the MOMS-PI study (@fettweis2019thev) that were made using the same 16S sequencing protocol.
We imputed the efficiencies of other species using taxonomic relationships (Methods).
Calibration then allowed us to explore the impact of bias on the measured trajectories of species proportions within women over the course of pregnancy as described in Section \@ref(case-studies).

@@

(ref:cap-calibration-example) **Fold changes can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet. The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which here is be known to be constant by construction). The 'Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species. The 'Calibrated (reference)' measurements are computed from abundances measured with the reference-species method, with _Lactobacillus crispatus_ used as the reference; that is, the true abundance of _L. crispatus_ is treated as known and used to infer the abundance of the remaining 6 species. Only samples that contain _L. crispatus_ are included.

@@

## Calibration using reference species

Obtaining control communities that span the taxa of interest is often not practical.
Moreover, the bias that is measured from controls may differ from that in the primary samples due to differences in cell state or sample preparation.
These problems may be overcome for the purpose of species-level absolute DA analysis by using the reference-species approach to absolute-density measurement that was described in Section \@ref(differential-abundance).
This *reference-species calibration* allows one to simultaneously account for changes in total density and the sample mean efficiency, obtaining accurate fold changes in absolute density from taxonomically biased, relative metagenomics measurements.

Recall that densities estimated by the reference-species approach have a constant fold error, provided that the fold error in the reference species is also constant (Equation \@ref(eq:density-ratio-error)).
Three categories of reference species are reviewed for this purpose in Appendix \@ref(review-absolute-methods), including why they might be expected to produce constant fold errors: housekeeping species assumed to have constant abundance, spike-in species added at known or constant abundance, and species whose abundance has been measured with a targeted method like species-specific qPCR.

To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species. 
Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).

## Choosing complementary total-density and metagenomics measurements

A variety of methods for measuring total community density have been paired with either amplicon or shotgun sequencing data for the purpose of conducting absolute DA analysis using the total-density approach described in Section \@ref(differential-abundance).
Each of these methods is subject to its own taxonomic bias: Cells from different species are likely to contribute more or less efficiently to the total abundance measurement.
Consider the popular method of assessing bacterial abundance with total-16S qPCR measurement.
Even an ideal qPCR protocol that perfectly amplifies all species remains a biased measurement of cell density, since the total 16S copies in the extracted DNA contributed by each species will be proportional to its (species-specific) lysis efficiency and 16S copy-number.
Empirical evidence supports this, as qPCR measurements of total 16S copies are rarely equivalent to total cell counts, even when multiple copy numbers are taken into account ()@lloyd2013meta).
16S qPCR is commonly paired with 16S amplicon sequencing, with which these sources of bias are shared, perhaps along with variation in primer binding and amplification efficiency.
Methods like flow cytometry that directly measure cell density lack these biases but likely have their own that are more likely to be orthogonal to the 16S sequencing measurement.
By extending the analysis of Section \@ref(differential-abundance) to include consistent taxonomic bias in the total-density measurement, we find that pairings of total-density and sequencing measurements that share large sources of taxonomic bias can lead to an offsetting of errors that reduces the error in fold-change measurement and DA inference (Appendix \@ref(total-density-bias)).
This finding suggests that methods of total-density measurement that are more accurate as measures of total cell density may actually perform worse than less accurate methods for the purposes of absolute DA inference.

We illustrate with a hypothetical example of two vaginal communities of equal density, but which are dominated either by _Lactobacillus iners_ or _Gardnerella vaginalis_.
Based on our earlier analysis in @mclaren2019cons, we estimate that _L. iners_ yields 10X more 16S copies per cell than _G. vaginalis_ in extracted DNA.
Ignoring other sources of bias, we expect relative efficiencies of _L. iners_:_G. vaginalis_ of 10 for both qPCR and 16S sequencing measurements.
Consequently, the _G. vaginalis_ community will yield a substantially lower qPCR measurement than the _L. iners_ community and we would incorrectly infer a decrease in total abundance.
At the same time, this decrease in the mean efficiency of the sequencing measurement would cause artificially high fold changes in the proportions of all species (Section \@ref(differential-abundance)).
When using qPCR and 16S sequencing measurements for community-based density estimation, these two opposing errors cancel, yielding accurate fold changes in species densities.

More generally, changes in the mean efficiency of the total-density measurement can offset those in the sequencing measurement when comparing log density of species across samples.
If the (log) efficiencies of the total and sequencing measurement are positively correlated across species, then the (log) mean efficiencies will tend to be positively correlated across samples, leading to a reduction in error in DA analysis relative to total-density methods whose efficiencies are uncorrelated with those of the sequencing measurement.
These observations suggest that qPCR of a marker gene may be the ideal pairing for amplicon sequencing measurements, despite being a poor measure of changes in total cell density.
Similarly, bulk DNA quantification may be an ideal pairing for shotgun sequencing, making it possible to account for variation in lysis efficiency and genome length.
These pairings even make it possible to account for error caused by variation among samples in the fraction of unclassified reads, which can form a major fraction of amplicon and shotgun data.
Importantly, maximizing the offsetting of errors requires thoughtful choices during bioinformatic analysis, perhaps eschewing the filtering and normalization steps used in many software packages and workflows (Appendix \@ref(total-density-bias)).

## Bias-sensitivity analysis

For experiments that have been conducted without calibration controls, it is still possible to investigate the likelihood that DA results can be explained by taxonomic bias by performing a computational bias-sensitivity analysis.

A bias-sensitivity analysis examines how sensitive the results of a given DA analysis are to the working assumption about the bias of a given protocol.
This working assumption is typically is that there is no bias, but could also be bias measured in a previous experiment or predicted from taxonomic features like 16S copy number.
A straightforward approach to bias-sensitivity analysis is to use computer simulation to re-analyze the MGS data across a set of simulated taxonomic biases.
First, multiple sets of efficiency vectors are randomly generated, each representing a possible quantification of taxonomic bias for the study.
Second, each efficiency vector is used to calibrate the original MGS measurements; each set of calibrated measurements represents the true community compositions under the given hypothesized bias. 
Finally, the DA analysis is re-run on each set of calibrated measurements and the distribution of results is compared graphically or with summary statistics.
A graphical summary of such an analysis being applied to results from the corncob frequentist DA- analysis tool 'corncob' are shown in SI Figure \@ref(fig:sensitivity-example).
The simulated efficiency vectors may be generated to different hypotheses about bias in the given system, such as its overall magnitude and correlation among phylogenetically-related species.
Thus the utility of such simulations will increase the more we learn about the properties of taxonomic bias in different systems.

An alternative approach to bias-sensitivity analysis would be to directly include the (unknown) taxonomic bias of the protocol in the statistical model used for DA analysis, as described for bias-aware meta-analysis below.
Further development of tools and workflows for performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.

@@

So far we have considered analysis of samples subject to the same taxonomic bias;
however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in samples from different studies.
This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses.
A potential solution is to perform a bias-aware meta-analysis that that explicitly accounts for the possibility of distinct taxonomic bias across studies.

Parametric microbiome meta-analysis models include sets of species-specific latent parameters to model "batch effects"—amorphous differences in the measurements of each study.
By configuring these models so that these latent parameters correspond to study-specific species efficiencies, we can improve their biological interpretability and gain the ability to directly assess whether disagreements in DA results across studies can be explained by differences in taxonomic bias.
To the extent that taxonomic bias truly is consistent within studies, meta-analyses that directly account for it can have greater statistical power to extract the truly consistent and inconsistent microbial patterns across studies.

>>>>

## Ratio-based relative DA analysis

Since bias creates constant fold errors (FEs) in the ratios among species, it can be countered by using relative DA methods that analyze FCs in these ratios.
A variety of ratio-based methods have been published under the name of Compositional Data Analysis (CoDA).
CoDA methods may more generally consider the ratios between products of multiple species, possibly raised to some exponent; a simple example is the geometric mean of several species.
The mathematical operations of multiplication and exponentiation maintain constant FEs.
Hence such products provide a method for aggregating species into a higher-level taxonomic units that maintains bias invariance when estimating FCs.
Ratio-based analyses also avoid the potential for misinterpreting changes in a species’ proportion that are driven by the sum-to-one constraint (@gloor2017micr).

We note two important caveats.
First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
In order to analyze FCs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
A common approach to enable the computation of FCs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
Unfortunately, this procedure violates bias invariance.
How bias interacts with more sophisticated approaches for handling zeros remains an important open question.
Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (@mclaren2019cons).
Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but may be harder to interpret.

If biological or technical considerations favor a DA method that is not bias invariant, it can be beneficial to also use a ratio-based method as a robustness check.
For example, @hevroni2020seas examined how the within-sample ranks of viral species changed from summer to winter.
Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.

@@

_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
Calibration can be extended to species not in the controls by imputing the efficiencies of the missing species using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties such as cell-wall structure.

Calibration makes it possible to correct for bias in relative and absolute DA methods that would otherwise not be bias invariant.
To demonstrate, we used calibration to improve the estimated FCs between samples in the absolute abundance of a species in the bacterial mixtures from @brooks2015thet.
We used the bias estimated from a single 7-species mixture to calibrate the MGS-derived proportions in all measured mixtures.
We obtained calibrated absolute abundances by multiplying the calibrated proportions by the known total abundance (Equation \@ref(eq:density-prop-meas)).
Calibration greatly increased the accuracy of the estimated FCs (Figure \@ref(fig:calibration-example), A vs. B).
The fungal and vaginal case studies of Section \@ref(case-studies) provide several practical examples of calibration in proportion-based DA analyses.

In studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome), calibration controls can be artificially constructed ('mock') communities.
In other cases, _natural community controls_ can be derived from aliquots of a homogenized microbiome sample.
Measurement of these controls across multiple experiments enables characterization of the _differential bias_ between experiments (@mclaren2019cons).
Calibration using differential bias can make results directly comparable across studies that used different.

@@

(ref:cap-calibration-example) **Fold changes can be calibrated using community controls or reference species.** The figure compares the performance of three methods for measuring fold changes in absolute cell density in cellular mock communities of 7 vaginal species, which were constructed and measured via 16S sequencing by @brooks2015thet. The 'Uncalibrated' fold changes are derived directly from uncalibrated individual abundance measurements, which equal the product of the species' proportion by the total density (which is constant by construction). The 'Calibrated (community)' measurements are computed from abundance measurements where the proportions are first corrected for the taxonomic bias that was estimated from a single sample that contained all 7 species. The 'Calibrated (reference)' measurements are computed from abundances measured with the reference-species method, with _Lactobacillus crispatus_ used as the reference; that is, the true abundance of _L. crispatus_ is treated as known and used to infer the abundance of the remaining 6 species. Only samples that contain _L. crispatus_ are included.

@@

## Choose absolute-abundance methods with more stable FEs

There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
The effect of bias on absolute DA analysis can be mitigated by choosing methods with more stable FEs, using either of two general approaches.

### Use complementary MGS and total-abundance measurements

A popular approach to measuring the absolute abundance of individual species is to multiply the species' proportions from MGS by a measurement of the abundance of the total community (Equation \@ref(eq:density-prop-meas)).
The resulting species abundances are affected by taxonomic bias in the MGS _and_ the total-abundance measurement (Equation \@ref(eq:density-prop-error)).
Section \@ref(differential-abundance) shows that the effect of both forms of bias on DA results can be reduced by choosing MGS and total-abundance methods that have similar taxonomic bias.

Consider the debate over whether flow cytometry or 16S qPCR are better methods for measuring total abundance for the purposes of normalizing 16S amplicon sequencing data (@galazzo2020howt, @jian2021comm).
Flow cytometry directly counts cells, whereas 16S qPCR measures the concentration of the 16S gene following DNA extraction.
Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate fold changes across samples.

More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
Optimal use of these pairings requires thoughtful choices during bioinformatics. For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements. Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
Future experiments should evaluate the extent to which the taxonomic bias of DNA-based total-abundance measurements is stable across samples and shared with the complementary MGS measurement.

### Normalize to a reference species

A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.

Previous studies have used species with a constant abundance as references: spike-ins, computationally-identified 'housekeeping species', and the host.
Researchers may naturally be concerned about bias between the reference species and the native species being normalized against it (see, for example, @harrison2021theq's recommendations for spike-in experiments).
Our results show that accurate FCs and DA results can be obtained so long as the relative efficiency of the focal species to the reference is consistent across samples.
This condition is much weaker than requiring that bias is small (the relative efficiency is close to 1) and so greatly expands the applicability of spike-ins and host normalization in microbiome experiments.

Section \@ref(absolute-abundance) further proposed an additional class of reference species for normalizing MGS measurements: Species whose abundance has been measured using a targeted method such as qPCR with species-specific primers.
To demonstrate the ability for reference calibration to improve fold changes, we treated the abundance of one species (_Lactobacillus crispatus_) in the @brooks2015thet mock community data as known, and used it to calibrate the abundances of all species. 
Doing so improved the resulting measurements of fold changes in density for all species (Figure \@ref(fig:calibration-example)).
Direct, targeted measurement of a reference species removes the need for it to have a constant abundance, thus further expanding the applicability of this reference-species normalization to experiments where spike-ins or housekeeping species are not viable options.
The targeted measurement need not itself be an unbiased measure of the reference species' abundance; so long as it has constant FE, the abundances of native species obtained by normalization will also have constant FEs.
In many ecosystems, there may not be a single universally-present species that can serve as a reference in all samples.
Sample coverage can be increased by making targeted measurements of multiple species or of larger taxonomic groups; see Appendix REF for details.

## Bias sensitivity analysis

Even if control measurements are not available, it is possible to computationally assess the sensitivity of a given DA result to taxonomic bias.
A _bias sensitivity analysis_ can be conducted by analyzing an MGS dataset multiple times under a range of possible taxonomic biases.
First, a large number of hypothetical biases are generated from a user-specified probability distribution.
Next, the DA analysis is re-run while using each simulated bias vector to calibrate the MGS measurements.
This approach is highly flexible, and can be used to investigate the bias sensitivity of any DA method.
Alternatively, existing DA methods can be extended to directly include the unknown taxonomic bias in their underlying statistical model, thereby providing DA estimates that inherently account for the added uncertainty in microbiome compositions due to the presence of unknown bias.
(@greenland2005mult compares these two approaches in a general context; @nixon2022asta applies the second approach to microbiome data in the absence of taxonomic bias.)

Interpreting a bias sensitivity analysis is complicated by the large percentage of zero counts in species-level microbiome profiles, as the results may strongly depend on how these zeros are handled in the calibration process.
Therefore it can be advantageous to jointly test the sensitivity of assumptions about zero-generating processes along with taxonomic bias, making statistical DA models that include bias and one or more zero-generating processes especially valuable.
The development of tools and workflows to facilitate bias sensitivity analysis may provide an efficient way to increase scientists’ ability to assess the reliability of microbiome results, both for differential abundance and microbiome analyses more generally.

@@

Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
These different biases can be explicitly accounted for in a _bias meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.
By estimating these 'nuisance parameters' along with the biological parameters of interest (such as species log-fold changes), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the 'batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
The ability to infer and adjust for the differential bias between studies can be increased by measuring one or more shared control samples alongside the experiment-specific samples.
